---
title: "Atlas integration"
author: "James Ashmore"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

## Motivation

## Setup

Setup chunk options:

```{r}
knitr::opts_chunk$set(
  autodep    = TRUE,
  cache      = TRUE,
  cache.lazy = FALSE,
  cache.path = "cache/02-atlas-integration.Rmd/",
  dev        = "png",
  error      = FALSE,
  message    = FALSE,
  warning    = FALSE,
  fig.align = "center",
  out.width = '100%'
)
```

Load Bioconductor packages:

```{r}
library(batchelor)
library(EnsDb.Mmusculus.v79)
library(MouseGastrulationData)
library(org.Mm.eg.db)
library(scater)
library(scran)
library(scuttle)
```

Load CRAN packages:

```{r}
library(pheatmap)
```

Source custom functions:

```{r}
source("code/cleanSCE.R")
source("code/cleanEmbryoAtlasData.R")
source("code/splitByCol.R")
```

## Data

### Test data

Read experiment objects:

```{r}
sce <- list(
  NMP_EHF = readRDS("data/NMP_EHF.rds"),
  LPMP_EHF = readRDS("data/LPMP_EHF.rds"),
  LPMP_LHF = readRDS("data/LPMP_LHF.rds")
)
```

Clean experiment objects:

```{r}
sce <- mapply(cleanSCE, x = sce, sample = names(sce))
```

### Reference data

Subset atlas sample metadata by embryonic stage:

```{r}
AtlasSampleMetadata <- subset(AtlasSampleMetadata, stage %in% c("E7.25", "E7.5", "E7.75", "E8.0", "E8.25"))
```

Retrieve processed samples from embryo atlas data:

```{r}
ref <- EmbryoAtlasData(type = "processed", samples = AtlasSampleMetadata$sample)
```

Clean processed samples from embryo atlas data:

```{r}
ref <- cleanEmbryoAtlasData(ref)
```

Split experiment data by the sample column:

```{r}
ref <- splitByCol(ref, ref$sample)
```

## Integration

### Data processing

Create list of all experiment data:

```{r}
all <- c(ref, sce)
```

Delete large objects and trigger garbage collection:

```{r}
rm(sce, ref); gc()
```

Subset data to common features:

```{r}
ids <- Reduce(intersect, lapply(all, rownames))

all <- lapply(all, "[", i = ids)
```

### Variance modelling

Model the per-gene variance:

```{r}
dec <- lapply(all, modelGeneVar)
```

Combine variance decompositions:

```{r}
dec <- do.call(combineVar, dec)
```

### Feature selection

Read blacklist genes and subset by category:

```{r}
nan <- read.csv("data/blacklist.csv")

nan <- subset(nan, CATEGORY %in% c("XIST", "Y", "MT", "CC", "RP"))
```

Select genes to ignore for feature selection:

```{r}
dec$blacklist <- FALSE

idx <- which(rownames(dec) %in% nan$ENSEMBL)

dec$blacklist[idx] <- TRUE
```

Identify highly variable genes:

```{r}
hvg <- subset(dec, blacklist == FALSE)

hvg <- getTopHVGs(hvg, var.field = "bio", var.threshold = 0, fdr.field = "p.value", fdr.threshold = 0.05)
```

### Batch normalization

Perform scaling normalization:

```{r}
all <- multiBatchNorm(all)
```

### Batch integration

Create sequential merge order:

```{r}
idx <- AtlasSampleMetadata

lvl <- c("E6.5", "E6.75", "E7.0", "mixed_gastrulation", "E7.25", "E7.5", "E7.75", "E8.0", "E8.25", "E8.5")

idx$stage <- factor(idx$stage, levels = lvl)

idx <- idx[order(idx$stage, idx$ncells, decreasing = TRUE), ]

idx <- c(idx$sample, "LPMP_LHF", "LPMP_EHF", "NMP_EHF")
```

Correct experiments using nearest neighbours:

```{r}
set.seed(1111001101)

mnn <- correctExperiments(
  all,
  subset.row = hvg,
  correct.all = TRUE,
  PARAM = FastMnnParam(merge.order = idx)
)
```

Delete large objects and trigger garbage collection:

```{r}
rm(all); gc()
```

Replace rowData slot:

```{r}
rowData(mnn) <- DataFrame(
  gene_id = rownames(mnn),
  gene_name = mapIds(EnsDb.Mmusculus.v79, keys = rownames(mnn), column = "GENENAME", keytype = "GENEID"),
  rotation = I(rowData(mnn)$rotation)
)
```

Create source annotation label:

```{r}
mnn$source <- "Atlas"

mnn$source[mnn$batch == "NMP_EHF"] <- "NMP_EHF"

mnn$source[mnn$batch == "LPMP_EHF"] <- "LPMP_EHF"

mnn$source[mnn$batch == "LPMP_LHF"] <- "LPMP_LHF"
```

Store combined variance model:

```{r}
metadata(mnn)$combineVar <- dec
```

Store highly variable genes:

```{r}
metadata(mnn)$getTopHVGs <- hvg
```

## Diagnostics

Check proportion of lost variance:

```{r}
var <- metadata(mnn)$merge.info$lost.var

col <- colorRampPalette(RColorBrewer::brewer.pal(n = 5, name = "Reds"))(100)

brk <- seq(0, max(var), length.out = 101)

pheatmap(var[, idx], color = col, display_numbers = TRUE, number_format = "%.2f", cluster_cols = FALSE, cluster_rows = FALSE, angle_col = 0)
```

Cluster on the corrected PCA to obtain a partitioning of the cells:

```{r}
snn <- buildSNNGraph(mnn, type = "jaccard", use.dimred = "corrected")

com <- igraph::cluster_louvain(snn)
```

Tabulate number of cells from multiple batches within each cluster:

```{r}
tab <- table(Cluster = com$membership, Batch = mnn$batch)

col <- colorRampPalette(RColorBrewer::brewer.pal(n = 5, name = "Blues"))(100)

brk <- seq(0, max(tab), length.out = 101)

pheatmap(tab[, idx], color = col, display_numbers = TRUE, number_format = "%d", cluster_cols = FALSE, cluster_rows = FALSE, angle_col = 0)
```

Perform TSNE on the corrected PCA matrix:

```{r}
set.seed(1101111001)

mnn <- runTSNE(mnn, dimred = "corrected", perplexity = 120) # same settings as reference analysis
```

Visualize the corrected PCA using a TSNE plot:

```{r fig.width = 9, fig.height = 9}
ggcells(mnn, aes(TSNE.1, TSNE.2, colour = celltype)) + 
  geom_point(size = 0.1, show.legend = FALSE) + 
  scale_colour_manual(values = EmbryoCelltypeColours, na.value = "#000000") + 
  labs(x = "TSNE 1", y = "TSNE 2") + 
  facet_wrap(~ sample) + 
  theme_bw() + 
  theme(aspect.ratio = 1)
```

Perform UMAP on the corrected PCA matrix:

```{r}
set.seed(0010111101)

mnn <- runUMAP(mnn, dimred = "corrected", n_neighbors = 20, min_dist = 0.7) # same settings as reference analysis
```

Visualize the corrected PCA using a UMAP plot:

```{r fig.width = 9, fig.height = 9}
ggcells(mnn, aes(UMAP.1, UMAP.2, colour = celltype)) + 
  geom_point(size = 0.1, show.legend = FALSE) + 
  scale_colour_manual(values = EmbryoCelltypeColours, na.value = "#000000") + 
  labs(x = "UMAP 1", y = "UMAP 2") + 
  facet_wrap(~ sample) + 
  theme_bw() + 
  theme(aspect.ratio = 1)
```

## Summary

### Output

Save integrated experiment object to disk:

```{r}
saveRDS(mnn, file = "output/02-atlas-integration.rds")
```

### Session

Print session information:

```{r}
sessionInfo()
```
