---
title: "Interactive data exploration"
author: "James Ashmore"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

## Setup

Load CRAN packages:

```{r}
library(rsconnect)
```

Include Bioconductor package repositories:

```{r}
options(repos = BiocManager::repositories())
```

Automatically answer yes to any user prompt:

```{r}
options(needs.promptUser = FALSE)
```

Delete existing shiny directory:

```{r}
unlink("shiny", recursive = TRUE)
```

Create shiny directory for interactive data exploration:

```{r}
dir.create("shiny", showWarnings = FALSE, mode = "0755")
```

## Applications {.tabset}

### Batch data

Create iSEE directory for batch data:

```{r}
dir.create("shiny/batch", showWarnings = FALSE, mode = "0755")
```

Copy batch data to iSEE directory:

```{r}
file.copy("output/01-batch-integration.rds", "shiny/batch/data.rds", overwrite = TRUE)
```

Write code to iSEE application file:

```{cat, engine.opts = list(file = "shiny/batch/app.R"), class.source = "r"}
# Load required packages
library(iSEE)
library(SingleCellExperiment)

# Load experiment object
sce <- readRDS("data.rds")

# Make feature names unique
rownames(sce) <- scuttle::uniquifyFeatureNames(rowData(sce)$gene_id, rowData(sce)$gene_name)

# Define column data variables
sce$Etv2 <- counts(sce)["Etv2", ] > 0
sce$Kdr <- counts(sce)["Kdr", ] > 0
sce$Runx1 <- counts(sce)["Runx1", ] > 0
sce$Ednrb <- counts(sce)["Ednrb", ] > 0
sce$Etv2_AND_Kdr <- sce$Etv2 & sce$Kdr
sce$Etv2_AND_Kdr_AND_Runx1 <- sce$Etv2_AND_Kdr & sce$Runx1
sce$Etv2_AND_Ednrb <- sce$Etv2 & sce$Ednrb

# Define column data colours
celltype_colours <- function(n) {
  c(
    "Allantois" = "#532C8A",
    "Anterior Primitive Streak" = "#C19F70",
    "Blood progenitors 1" = "#D9DECF",
    "Blood progenitors 2" = "#C9A997",
    "Cardiomyocytes" = "#B51D8D",
    "Caudal epiblast" = "#9E6762",
    "Caudal Mesoderm" = "#3F84AA",
    "Caudal neurectoderm" = "#354E23",
    "Def. endoderm" = "#F397C0",
    "Endothelium" = "#FF891C",
    "Epiblast" = "#635547",
    "Erythroid1" = "#C72228",
    "Erythroid2" = "#F79083",
    "Erythroid3" = "#EF4E22",
    "ExE ectoderm" = "#989898",
    "ExE endoderm" = "#7F6874",
    "ExE mesoderm" = "#8870AD",
    "Forebrain/Midbrain/Hindbrain" = "#647A4F",
    "Gut" = "#EF5A9D",
    "Haematoendothelial progenitors" = "#FBBE92",
    "Intermediate mesoderm" = "#139992",
    "Mesenchyme" = "#CC7818",
    "Mixed mesoderm" = "#DFCDE4",
    "Nascent mesoderm" = "#C594BF",
    "Neural crest" = "#C3C388",
    "NMP" = "#8EC792",
    "Notochord" = "#0F4A9C",
    "Paraxial mesoderm" = "#8DB5CE",
    "Parietal endoderm" = "#1A1A1A",
    "PGC" = "#FACB12",
    "Pharyngeal mesoderm" = "#C9EBFB",
    "Primitive Streak" = "#DABE99",
    "Rostral neurectoderm" = "#65A83E",
    "Somitic mesoderm" = "#005579",
    "Spinal cord" = "#CDE088",
    "Surface ectoderm" = "#F7F79E",
    "Visceral endoderm" = "#F6BFCB"
  )
}

boolean_colours <- function(n) {
  c("TRUE" = "#E15759", "FALSE" = "#BAB0AC")
}

# Create iSEE object
ecm <- ExperimentColorMap(colData = list(celltype = celltype_colours), global_discrete = boolean_colours)
iSEE(sce, appTitle = "LPMP_NMP", colormap = ecm)
```

Deploy the iSEE application to the shiny server:

```{r}
deployApp(appDir = "shiny/batch", appName = "LPMP_NMP", appTitle = "LPMP_NMP", launch.browser = FALSE, forceUpdate = TRUE)
```

Allocate more memory to the iSEE application:

```{r}
configureApp(appName = "LPMP_NMP", appDir = "shiny/batch", size = "xxlarge")
```

### Atlas data

### Human data

Create iSEE directory for human data:

```{r}
dir.create("shiny/human", showWarnings = FALSE, mode = "0755")
```

Copy human data to iSEE directory:

```{r}
file.copy("output/03-human-integration.rds", "shiny/human/data.rds", overwrite = TRUE)
```

Write code to iSEE application file:

```{cat, engine.opts = list(file = "shiny/human/app.R"), class.source = "r"}
# Load required packages
library(iSEE)
library(SingleCellExperiment)
library(BiocSingular)

# Load experiment object
sce <- readRDS("data.rds")

# Make feature names unique
rownames(sce) <- scuttle::uniquifyFeatureNames(rowData(sce)$Mouse_ID, rowData(sce)$Mouse_Symbol)

# Define column data variables
sce$Etv2 <- counts(sce)["Etv2", ] > 0
sce$Kdr <- counts(sce)["Kdr", ] > 0
sce$Runx1 <- counts(sce)["Runx1", ] > 0
sce$Ednrb <- counts(sce)["Ednrb", ] > 0
sce$Etv2_AND_Kdr <- sce$Etv2 & sce$Kdr
sce$Etv2_AND_Kdr_AND_Runx1 <- sce$Etv2_AND_Kdr & sce$Runx1
sce$Etv2_AND_Ednrb <- sce$Etv2 & sce$Ednrb

# Define column data colours
sample_colours <- function(n) {
  c(
    "Human" = "#BAB0AC",
    "NMP_EHF" = "#4E79A7",
    "LPMP_EHF" = "#E15759",
    "LPMP_LHF" = "#EDC948"
  )
}

cluster_colours <- function(n) {
  c(
    "Epiblast" = "#3D7AB0",
    "Ectoderm" = "#4FC0CB",
    "Primitive Streak" = "#A6BEDF",
    "Nascent Mesoderm" = "#C67DB4",
    "Axial Mesoderm" = "#7B534D",
    "Emergent Mesoderm" = "#ABB973",
    "Endoderm" = "#F28941",
    "Advanced Mesoderm" = "#4F9E74",
    "YS Mesoderm" = "#FAB781",
    "Hemogenic Endothelial Progenitors" = "#6A59A6",
    "Erythrocytes" = "#C5433E"
  )
}

origin_colours <- function(n) {
  c(
    "Caudal" = "#4DAF4A",
    "Rostral" = "#377EB8",
    "Yolk Sac" = "#FF7F00"
  )
}

boolean_colours <- function(n) {
  c(
    "TRUE" = "#E15759",
    "FALSE" = "#BAB0AC"
  )
}

# Create iSEE object
ecm <- ExperimentColorMap(colData = list(Sample = sample_colours, Cluster = cluster_colours, Origin = origin_colours), global_discrete = boolean_colours)
iSEE(sce, appTitle = "LPMP_NMP_HUMAN", colormap = ecm)
```

Deploy the iSEE application to the shiny server:

```{r}
deployApp(appDir = "shiny/human", appName = "LPMP_NMP_HUMAN", appTitle = "LPMP_NMP_HUMAN", launch.browser = FALSE, forceUpdate = TRUE)
```

Allocate more memory to the iSEE application:

```{r}
configureApp(appName = "LPMP_NMP_HUMAN", appDir = "shiny/human", size = "xxlarge")
```

## Summary

### Session

Print session information:

```{r}
sessionInfo()
```
